Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-29 paper_jasa.tex
edited as suggested by Hal

\noindent where $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.
$W$ is a design matrix of covariates corresponding to specifics of lineup $\ell$ and subject $i$, and $\delta$ is the vector of corresponding parameters. Covariates could include  demographic information of individuals, such as age, gender, education level etc.,  as well lineup-specific elements, e.g. effect size or difficulty level.
%
$Z_{\ell i}$,  $1 \le i \le K$, $1 \le \ell \le L$,  is a design matrix corresponding to random effects specific to individual $i$ and lineup $\ell$; and 
$\tau$  is a vector of independent normally distributed random variables $\tau_{\ell i}$ with  variance matrix $\sigma_\tau I_{KL \times KL}$. $\tau$ will usually include a component incorporating an individual's ability or skill to evaluate lineups. 
Note that $t_{\ell i }$ usually only includes a partial interaction; for a full interaction of subjects' skills and lineup-specific difficulty we would need replicates of the same subject evaluating the same lineup, which in practice is not feasible without loosing independence.

@@

In this setting, there are many established graphics that are used to evaluate and diagnose the fit of a regression model (e.g. \citet{cook:99}). Table \ref{tbl:stat_multiple} lists several common hypotheses related to the regression setting, and commonly used statistical plots that might be used as corresponding visual test statistics.
%These plots should be familiar, because they are all in common use. 
For example, to examine the effect of variable $X_j$ on $Y$, we would plot $Y$ against $X_j$ (cases 1-4). To assess whether the linear assumption is appropriate we would draw a plot of residuals against fitted values (case 5). For the purposes of comparing visual against conventional inference, we focus on cases 2 and 3, with a continuous and categorical explanatory variable respectively.

@@

\subsection{Continuous covariate}

@@

\begin{figure}[p]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_continuous.pdf}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope?}
       \label{fig:test_continuous}
\end{figure}

\subsection{Contaminated data}

@@

\begin{figure}[p]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_contaminated.pdf}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous but the inclusion of some contamination with the data spoils the normality assumption of error structure. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope?}
       \label{fig:test_contaminated}
\end{figure}

>>>>

\noindent where $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.
$W$ is a design matrix of covariates corresponding to specifics of lineup $\ell$ and subject $i$, and $\delta$ is the vector of corresponding parameters. Covariates could include  demographic information of individuals, such as age, gender, education level etc.,  as well lineup-specific elements, e.g. effect size or difficulty level.
%
$Z_{\ell i}$,  $1 \le i \le K$, $1 \le \ell \le L$,  is a design matrix corresponding to random effects specific to individual $i$ and lineup $\ell$; and 
$\tau$  is a vector of independent normally distributed random variables $\tau_{\ell i}$ with  variance matrix $\sigma_\tau I_{KL \times KL}$. $\tau$ will usually include a component incorporating an individual's ability or skill to evaluate lineups. 
Note that $\tau_{\ell i }$ usually only includes a partial interaction; for a full interaction of subjects' skills and lineup-specific difficulty we would need replicates of the same subject evaluating the same lineup, which in practice is not feasible without loosing independence.

@@

In this setting, there are many established graphics that are used to evaluate and diagnose the fit of a regression model (e.g. \citet{cook:99}). Table \ref{tbl:stat_multiple} lists several common hypotheses related to the regression setting, and commonly used statistical plots that might be used as corresponding visual test statistics.
%These plots should be familiar, because they are all in common use. 
For example, to examine the effect of variable $X_j$ on $Y$, we would plot $Y$ against $X_j$ \blue{(cases 1-4)}. To assess whether the linear assumption is appropriate we would draw a plot of residuals against fitted values (case 5). For the purposes of comparing visual against conventional inference, we focus on cases 2 and 3, with a continuous and categorical explanatory variable respectively.

@@

\subsection{Continuous covariate}  \label{sec:continuous}

@@

\begin{figure}[p]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_continuous.pdf}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope? (The position of the actual data plot is provided in Section \ref{sec:continuous}.)}
       \label{fig:test_continuous}
\end{figure}

\subsection{Contaminated data} \label{sec:contamination}

@@

\begin{figure}[p]
%\begin{figurehere}
   \centering
       \includegraphics[width=0.95\textwidth]{lineup_contaminated.pdf}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous but the inclusion of some contamination with the data spoils the normality assumption of error structure. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope? \blue{(The position of the actual data plot is provided in Section \ref{sec:contamination}.)}}
       \label{fig:test_contaminated}
\end{figure}
